Nonlinear Tidal Dissipation in White Dwarfs

### 2019-01-18 proposal/NDSEG_2017/NDSEG_proposal.tex
reorg proposals

\def\Snospace~{\S{}} % hack to remove the space left after autorefs
\renewcommand*{\sectionautorefname}{\Snospace}
\renewcommand*{\appendixautorefname}{\Snospace}
\renewcommand*{\figureautorefname}{Fig.}
\renewcommand*{\equationautorefname}{Eq.}
\renewcommand*{\tableautorefname}{Tab.}

\onehalfspacing

@@

White dwarfs (WDs) are luminous, long-lasting products of late-stage stellar
evolution that are held up by electron degeneracy pressure, a quantum mechanical
effect arising from subjecting electrons to immense pressures. They are some of
the densest objects in the known universe, behind only neutron stars (NS) and
black holes (BH), fitting a solar mass into an Earth-sized sphere. Furthermore,
they exhibit various chemical compositions, each of which is subject to
different physics, and are found in a variety of interesting systems. As such,
WDs are a unique and diverse window into matter under extreme conditions.

WDs are commonly found in binary orbits with companion objects, in which the two
orbit their center of mass under their mutual gravitational attraction. The
companion object ranges from another WD to a supermassive BH (SMBH),
hypothesized to be at the center of galaxies and have mass $10^5$--$10^6$ solar
masses. All of these binary systems are very important to astrophysics. WD-WD
binaries are most important for being thought to generate \emph{Type Ia
supernovae} in which one of the two WDs reaches a critical mass through
accreting surrounding matter and collapses under its own gravity, producing a
shock wave that tears the WD apart and releases immense amounts of energy. Type
Ia supernovae are unique in astrophysics for their near-constancy throughout
space and time, consistently occuring at the WD critical mass, and have been
used to discover numerous cosmological phenomena such as dark energy.

WD-BH systems are also interesting subjects of study. Studies indicate that
as WDs orbit the extreme gravity surrounding BHs, they will produce observable
flares induced by gravitational tidal forces\cite{flares}. A WD orbiting a SMBH
would produce gravitational waves, waves carried by distortions in space-time
and generated according to Einstein's general relativity by accelerating massive
objects, that are expected to be detected by the space-based \emph{Laser
Interferometer Space Antenna (LISA)} when deployed\cite{lisa}. Gravitational
wave astronomy is an increasingly exciting field as the \emph{Laser
Interferometer Gravitational-Wave Observatory (LIGO)} continues to make
detections since its first in late 2015. As gravitational wave astronomy relies
on accurate predictions of the expected signals, it is important to build as
accurate models as possible before observation runs begin.

@@

The excitation of \emph{internal gravity waves} in the WD by the tidal forces of
the companion is an effect seen in all of the aforementioned systems. Internal
gravity waves, not to be confused with the gravitational waves discussed above,
are internal displacements in the WD that oscillate and propagate due to a
restoring buoyancy force. As these waves propagate outward from where they are
excited, they are expected to grow in amplitude until they break, as do ocean
waves on a shore, and deposit both energy and angular momentum inside the WD\@.

Previous work predicts that this dissipation mechanism can generate
significantly more energy than thermal radiation from the WD surface alone and
are thus a significant contribution to the WD energy budget\cite{fullerII}. The
exact radial dissipation profile is of enormous interest since it is both
sensitive to WD properties and can produce drastically different observable
outcomes. One proposed outcome is a \emph{tidal nova}, in which heating in the
WD's degenerate hydrogen layer is significant enough to trigger runaway nuclear
fusion and a significant, observable explosion\cite{tidal_novae}. Understanding
whether such phenomena occur requires understanding how energy is distributed
internally inside a WD\@.

Internal gravity wave breaking is a nonlinear hydrodynamic phenomenon. Such
phenomena are known to require numerical simulation to study. It is therefore
paramount to begin numerical study to build dissipation models inside WDs to
characterise what phenomena can be observed in which WD models and
characteristics.

@@

\textbf{We propose to study tidal dissipation via nonlinear gravity wave
breaking in white dwarfs via numerical simulation.}

Our research will consist of computing for various WD models and compositions
the energy and angular momentum dissipation profiles inside WDs. Once such
profiles are obtained, we intend to add these profiles to existing stellar
evolution codes to study the dynamical effects of tidal dissipation. We will
attempt to find a compact or even analytical representation of our numerical
work to greatly simplify and accelerate such integration. Finally, we will make
observational predictions with our results and compare our predictions with
existing observed WDs. Any software and results would be made in accordance with
best practice and publicly available for the entire community.

\subsection{Qualifications}

The proposing researcher (YS) and his adviser (DL) are uniquely qualified to
pursue such studies. YS has ample research experience in numerical simulation
and is continuing to pursue academic study in hydrodynamics, turbulence and
computation. He also completed a double Bachelors degree in physics and computer
science from the California Institute of Technology. Finally, he has worked in
the software industry for a year and continues to follow discussions of best
practice far ahead of those in academia and apply them to his own work.

DL is a co-author on much of the latest literature on WD binaries and is an
expert on many related subjects such as fluid dynamics in extreme matter and
studies of other compact object systems. Cornell University also houses Saul
Teukolsky's research group, one of the pre-eminent numerical relativity groups
in the world, which will be frequently consulted for numerical best practices.

In terms of software, YS proposes to only implement what is not currently
available in the community. Current literature indicates that the spectral
hydrodynamic code Dedalus is able to handle instability-driven turbulence with
high accuracy and speed, and our preliminary work seems to indicate that Dedalus
will be sufficiently accurate to simulate wave breaking
dissipation\cite{dedalus}. Modules for Experiments in Stellar Astrophysics
(MESA) is a proven stellar evolution code that will likely need only be extended
to incorporate the discovered dissipation models\cite{MESA}. Nevertheless, both
codes will require adaptation to be applied to the present problem (YS has
already contributed an improvement to Dedalus), a task for which YS is
well-equiped.

Should no existing software prove adequate, we are more than able and willing to
develop a new code and contribute it to the community. Notably, no
GPU-accelerated spectral hydrodynamic code has yet been released, which would be
of great value to researchers, and YS is well-versed in GPU programming.
Nevertheless, in the interest of the proposed science objectives codes will be
developed on an as-needed basis.

@@

A robust pipeline coupling nonlinear wave breaking to system evolution
would be of interest spanning all fields of physics. Barring BHs, all
astrophysical objects exhibit internal waves that can break as they propagate
towards the surface. Notable present interests in the astrophysical community
include tidal disruption, planet formation and stars. Even in planetary and
atmospheric sciences, wave breaking in atmospheres would share many common
techniques with the proposed research.

The proposed research is extremely relevant to Department of Defense (DoD)
fields of interest. WDs and its energetic phenomena we propose to study e.g.\
novae of all sorts are natural laboratories of extreme physics that have
direct consequences for understanding of plasmas and other technologies integral
to propulsion in aeronautics. A theory of nonlinear wave breaking would have
far-reaching consequences in turbulence study and atmospheric sciences. Finally,
the observation of our predictions could shed light on atmospheric optics,
whether as a use for calibration or understanding contamination effects.

>>>>

\def\Snospace~{\S{}} % hack to remove the space left after autorefs
\renewcommand*{\sectionautorefname}{\Snospace}
\renewcommand*{\subsectionautorefname}{\Snospace}
\renewcommand*{\appendixautorefname}{\Snospace}
\renewcommand*{\figureautorefname}{Fig.}
\renewcommand*{\equationautorefname}{Eq.}
\renewcommand*{\tableautorefname}{Tab.}

\setstretch{1.2}

@@

White dwarfs (WDs) are remnants of stellar evolution for stars with mass
$\lesssim 8 M_{sun}$. They are some of the densest objects in the universe,
fitting a solar mass into an Earth-sized sphere. They are supported against
gravity by electron degeneracy pressure, a quantum mechanical effect arising from
subjecting electrons to immense pressures Furthermore, they exhibit various
chemical compositions, each of which is subject to different physics, and are
found in a variety of interesting systems. As such, WDs are a unique and diverse
window into matter under extreme conditions.

WDs are commonly found in binaries in which two objects orbit their center of
mass under mutual gravitational attraction. The companion object ranges
from another WD to a supermassive BH (SMBH), hypothesized to be at the center of
galaxies and have mass $\gtrsim 10^5$ solar masses. All of these binary systems
are very important to astrophysics. WD-WD binaries are most important for being
thought to generate \emph{Type Ia supernovae} in which two WDs merge and
initiate runaway thermonuclear fusion that powers an explosion, releasing
immense amounts of energy. Because Type Ia supernovae (SNe Ia) are highly
luminous, can be seen at large distances and are very consistent, they have been
used as a ``standard candle'' to probe the expansion rate of the universe (e.g.\
observations of SNe Ia provided the first evidence of the existence of dark
energy in the universe).

WD-BH systems are also interesting subjects of study. Recent works indicate that
as WDs plunge close to BHs, they will produce observable flares induced by
gravitational tidal forces\cite{flares}. A WD orbiting a SMBH would also produce
gravitational waves, waves of space-time warping as predicted by Einstein's
theory of general relativity, that are expected to be detected by the
space-based \emph{Laser Interferometer Space Antenna (LISA)} when
deployed\cite{lisa}. Gravitational wave astronomy is an increasingly exciting
field as the \emph{Laser Interferometer Gravitational-Wave Observatory (LIGO)}
continues to make progress since its first detection in late 2015. As
gravitational wave astronomy relies on accurate predictions of the expected
signals, it is important to build as accurate models as possible before
observation runs begin.

@@

The excitation of \emph{internal gravity waves} in the WD by the tidal forces of
the companion is an effect exhibited in all of the aforementioned systems.
Internal gravity waves, not to be confused with the gravitational waves above,
are internal displacements in the WD fluid that oscillate and propagate due to a
restoring buoyancy force. These tidally excited gravity waves are analogous to
tides on Earth raised by the Moon and the Sun, except that since WDs do not have
sharp surfaces these waves are internal to the WD\@. As these waves propagate
outward from where they are excited, they are expected to grow in amplitude
until they break, as do ocean waves on a shore, and deposit both energy and
angular momentum in the WD envelope.

Previous work predicts that this dissipation mechanism can generate
significantly more energy than thermal radiation from the WD surface alone and
are thus a significant contribution to the WD energy budget\cite{fullerII}. The
exact radial dissipation profile is of interest since it both is sensitive to WD
properties and can produce vastly different observable outcomes. One proposed
outcome is a \emph{tidal nova}, in which heating in the WD's degenerate hydrogen
layer is sufficient to trigger runaway nuclear fusion and an observable
surface explosion\cite{tidal_novae}. Understanding whether such phenomena occur
requires understanding how energy is distributed internally inside a WD\@.

Internal gravity wave breaking is a nonlinear hydrodynamic phenomenon. Such
phenomena are known to require numerical simulation to study. It is therefore
paramount to begin numerical study to build dissipation models inside WDs to
characterize what phenomena can be observed in different WD models.

@@

\textbf{We propose to study tidal dissipation via nonlinear gravity wave
breaking in white dwarfs using numerical simulation.}

Our research will initially consist of numerically computing for various WD
models and compositions the energy and angular momentum dissipation profiles
inside WDs. Once such profiles are obtained, we intend to add these profiles to
existing stellar evolution codes to study the dynamical effects of tidal
dissipation. We will attempt to find a compact or even analytical representation
of our numerical work to greatly simplify and accelerate such integration.
Finally, we will make observational predictions with our results and compare our
predictions with existing observed WDs in binaries. Any software and results
would be made in accordance with best practice and publicly available for the
scientific community.

\subsection{Numerical Approach}

Much of our proposed research will rely on numerical solutions to fluid
dynamical equations. Guided by current literature, we have begun with the
Dedalus numerical solver, a modern spectral solver that integrates the fluid
equations both quickly and to high accuracy\cite{dedalus}. Dedalus has been
shown in current literature to describe turbulence and fluid instabilities well,
and our preliminary work supports this conclusion.

For our problem, we will first adapt Dedalus to study gravity waves in the
stratified atmosphere/envelope of WDs and later consider global (spherical) WD
models. Should Dedalus prove inadequate, we are able (\autoref{ss:quals}) and
willing to contribute a new hydrodynamic code to the community. Notably, no
GPU-accelerated spectral hydrodynamic code has yet been released, which would be
of great value to researchers. Nevertheless, in the interest of the proposed
science objectives codes will be developed on an as-needed basis.

After characterizing tidal dissipation, we intend to apply our models to WD
evolution under tidal heating. Modules for Experiments in Stellar Astrophysics
(MESA) is a proven stellar evolution code that would require comparatively
little adaptation to incorporate tidal heating in studying WD
evolution\cite{MESA}. A released extension to MESA containing our work would be
the most effective way to release our work to prospective users.

\subsection{Qualifications}\label{ss:quals}

The proposing researcher (YS) and his adviser (DL) are uniquely qualified to
pursue such studies. YS has completed a double Bachelors degree in physics and
computer science from the California Institute of Technology. He also has ample
research experience in numerical simulation and is continuing to pursue academic
study in hydrodynamics, turbulence and computation. Finally, he has worked in
the software industry for a year and continues to follow discussions of best
practice far ahead of those in academia and apply them to his own work.

DL is a co-author of a number of recent papers on WD binaries and is an
expert on many related subjects such as fluid dynamics in extreme matter and
studies of other compact object systems. Cornell University also houses Saul
Teukolsky's research group, one of the pre-eminent numerical relativity groups
in the world, which will be frequently consulted for numerical best practices.

@@

Although our proposed research deals with astronomical objects and phenomena
(WDs and BHs), the tools we use and develop in our research
have applications in many areas of physical sciences and engineering. As
discussed above, through our research we will develop extensive expertise in
nonlinear hydrodynamics (internal gravity waves), radiation physics (radiative
transfer and diffusion of heat inside WD), nuclear physics (tidal heating
leading to runaway fusion) and numerical computation. Such a research project
thus contributes to many science and engineering fields and equips me to tackle
an exceptionally large variety of problems.

The proposed research is extremely relevant to Department of Defense (DoD)
fields of interest. WDs and its energetic phenomena we propose to study (e.g.\
novae) of all sorts are natural laboratories of extreme physics that have
direct consequences for understanding of plasmas and other technologies integral
to propulsion in aeronautics. A theory of nonlinear wave breaking would have
far-reaching consequences in turbulence study and atmospheric sciences. Finally,
the observation of our predictions could shed light on atmospheric optics,
whether as a use for calibration or understanding contamination effects.
